Supplementary Materials for \" COATi: statistical pairwise alignment of protein coding sequences\

### 2023-05-15 results_discussion.tex
Add suppl method kaks paragraph

After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 6048 sequence pairs that contained gaps identified by at least one aligner, and 7719 ungapped sequence pairs.
We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\citembe{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\citembe{ka_ks_li_1993}.
The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 0 representing a perfect result.

@@

COATi, using the codon-triplet-ecm model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.06 \cdot 10^{-79}$ according to the one-tailed Wilcoxon signed rank test.
In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
It obtained better results compared to a wide variety of alignment strategies.
Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.

>>>>

After downloading, we removed 2232 sequences longer than 6000 nucleotides, identified 6048 sequence pairs that contained gaps identified by at least one aligner, and 7719 ungapped sequence pairs.
We then randomly introduced gap patterns extracted from all five methods into the ungapped sequence pairs to generate the benchmark alignments.
Alignment accuracy was measured using the distance metric $d_{seq}$
\citembe{metrics_blackburne_whelan_2011} between simulated and inferred
alignments.
In addition, accuracy of positive and negative selection was calculated
using the $F_1$ score by estimating $k_s$ and $k_a$ statistics
\citembe{ka_ks_li_1993} (Supplementary Methods).
The $F_1$ score evaluates the accuracy of a model by assigning equal importance to precision and recall, and ranges between 0 and 1, with a score of 0 representing a perfect result.

@@

COATi, using the codon-triplet-mg model, was significantly more accurate (lower $d_{seq}$) at inferring simulated alignments compared to other methods; all p-values were less than $2.06 \cdot 10^{-79}$ according to the one-tailed Wilcoxon signed rank test (Supplementary Materials Figure 1).
In addition, COATi produced more perfect alignments, less imperfect alignments, and more accurately retrieved events of positive selection (Table \ref{table:comp}).
It obtained better results compared to a wide variety of alignment strategies.
Clustal$\Omega$, performing a common approach of aligning via amino acid translations, obtained the highest average alignment error and had difficulties retrieving positive selection.
MACSE, which allows frameshifts, is also based on an amino acid model and obtained similar results to the DNA-based MAFFT.
PRANK, using a codon model, had a similar average alignment error to MACSE and MAFFT but had issues recovering the simulated alignments.

### 2023-05-15 supplementary_materials.Rmd
Add suppl method kaks paragraph

---
title: "Supplementary Materials"
#author: "Juan J. García Mesa"
#date: "`r Sys.Date()`"
geometry: margin=1.8cm
output:
    pdf_document:
        fig_caption: yes
---

@@

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg"))
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm"))
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg"))
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm"))
```

@@

>>>>

---
title: "Supplementary Materials for \" COATi: statistical pairwise alignment of protein coding sequences\""
author: Juan J. Garcia Mesa, Ziqi Zhu, Reed A. Cartwright
output:
    pdf_document:
        fig_caption: yes
bibliography: alignpair_letter.bib
---

@@

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_mg, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/tri-mg/plot_distance.csv", "tri-mg"))
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-tri-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-triplet-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_tri_ecm, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/tri-ecm/plot_distance.csv", "tri-ecm"))
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-mg} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-mg and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_mg, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/mar-mg/plot_distance.csv", "mar-mg"))
```

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.align = 'center', fig.cap = paste0("\\label{fig:dseq-mar-ecm} Comparison of log10-transformed $d_{seq}$ data with pseudocounts between COATi codon-marginal-ecm and PRANK, MAFFT, ClustalOmega, and MACSE. COATi was significantly more accurate than other aligners; all p-values were $\\leq ", signif(pval_mar_ecm, digits = 3), "$.")}
library(knitr)
source("supplementary_materials/scripts/plot_dseq.R")
suppressWarnings(plot_dseq_main("supplementary_data/mar-ecm/plot_distance.csv", "mar-ecm"))
```

@@

# Supplementary Methods

Ks and Ka represent the number of substitutions per synonymous and non-synonymous sites. The ratio of nonsynonymous (Ka) to synonymous (Ks) nucleotide substitution rates indicates the selective pressures acting on genes. If the ratio is significantly greater than 1, it suggests positive selective pressure, meaning that nonsynonymous substitutions occur more frequently than synonymous substitutions. A ratio around 1 can indicate either neutral evolution at the protein level or a mixture of positive and negative selective pressures. If the ratio is less than 1, it indicates a pressure to maintain protein sequence, known as purifying selection.
Ks and Ka are calculated using the R package seqinr v.4.2-30 [@seqinr].

# References

<div id="refs"></div>
